---
title: "Nonlinear Manifold Learning: ISOMAP and LLE"
author: "Alex Young"
date: "10/29/2019"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE, message = FALSE)
set.seed(1) # for reproducibility
require("dimRed")
require("energy")
require("scatterplot3d")
require("rgl")
require("plotly")
# 
myColorRamp <- function(colors, values) {
    v <- (values - min(values))/diff(range(values))
    x <- colorRamp(colors)(v)
    rgb(x[,1], x[,2], x[,3], maxColorValue = 255)
}

# generate helix data
t <- sort(runif(500, min = 0, max = 15))
helix <- matrix(NA,nrow = length(t), ncol = 3)
for (n in 1:length(t)){
  helix[n,] <- c( cos(t[n]), sin(t[n]) , t[n])
}

# generate Swiss-shaped data
S <- rep(0,2000)
Swiss <- matrix(NA, nrow = 2000, ncol = 3)
for( n in 1:2000){
    s <- runif(1, min = 3*pi/2, max = 9*pi/2)
    t <- runif(1, min = 0,      max = 15)
    S[n] <- s
    Swiss[n, ] <- c( s*cos(s), t, s*sin(s) )
}

# generate washer in 2d 
N <- 1e5
washer <- matrix(NA, nrow =N, ncol = 2)
for (n in 1:N){
  r <- runif(1,min=5,max = 10); 
  a = runif(1,min = 0, max = pi); 
  washer[n,] <- r*c(cos(a),sin(a)) + c(20,0)
  }

# generate folded washer in 3D
washer3 <- cbind(washer, washer[,2]^2)

# generate rolled washer in 3D
washer.swiss <- washer3
for (n in 1:dim(washer3)[1] ){
  washer.swiss[n,] <- c( washer[n,1]*cos(washer[n,1]), washer[n,2], washer[n,1]*sin(washer[n,1] )) 
}

```




#  {.tabset}

## Data and 3D Scatterplots {.tabset}

### Helix
```{r, echo = FALSE}
 scatterplot3d(helix, color = myColorRamp(c("red","blue"), helix[,3] ),
               xlab = expression(x[1]), ylab = expression(x[2]), zlab = expression(x[3]))

```

### Swiss Roll
```{r, echo = FALSE}
scatterplot3d(Swiss, color = myColorRamp(c("red","purple","blue","green","yellow"), S ),
               xlab = expression(x[1]), ylab = expression(x[2]), zlab = expression(x[3]))
```


### Washer
```{r, echo = FALSE}
plot(washer, xlab = '', ylab = '',
     main = "Washer in 2D", col = myColorRamp(c("red","purple","blue","green","yellow"), washer[,1] )
)
```

### Folded Washer
```{r}
scatterplot3d(washer3,
              color = myColorRamp(c("red","purple","blue","green","yellow"), washer[,1] ),
               xlab = expression(x[1]), ylab = expression(x[2]), zlab = expression(x[3]),
              main = "Folded Washer")
```

### Rolled Washer

```{r}
scatterplot3d(washer.swiss,
              color = myColorRamp(c("red","purple","blue","green","yellow"), washer[,1]),
              xlab = expression(x[1]), ylab = expression(x[2]), zlab = expression(x[3]),
              main = "Rolled Washer",
              cex.symbols  = 0.1
)
```

## ISOMAP {.tabset}

Below, we show the manifold learning of the given manifolds using ISOMAP with $K = 20$ nearest neighbors.

### Helix

```{r}
par(mfrow = c(1,2) )
scatterplot3d(helix, color = myColorRamp(c("red","blue"), helix[,3]),
               xlab = expression(x[1]), ylab = expression(x[2]), zlab = expression(x[3]) )
test <- embed(helix, "Isomap", .mute = c("message", "output"), ndim =1)
distance_correlation(test)
plot(test@data@data, rep(0,dim(helix)[1]),
     xlab = "1st Dimension", ylab = "", yaxt = 'n',
     col = myColorRamp(c("red","blue"), helix[,3]) )
```

### Swiss Roll

```{r, message= FALSE}
par(mfrow = c(1,2))
scatterplot3d(Swiss, color = myColorRamp(c("red","purple","blue","green","yellow"), S ),
               xlab = expression(x[1]), ylab = expression(x[2]), zlab = expression(x[3]))
plot(embed(Swiss, "Isomap", .mute = c("message", "output"), ndim =2, knn = 20)@data@data,
     xlab = "1st Dimension", ylab = "2nd Dimension",
     col = myColorRamp(c("red","purple","blue","green","yellow"), S)  
     )
```

### Folded Washer

```{r}
N <- 2000
par(mfrow = c(1,2))
scatterplot3d(washer3[1:N,], color = myColorRamp(c("red","purple","blue","green","yellow"), washer[1:N,1] ),
               xlab = expression(x[1]), ylab = expression(x[2]), zlab = expression(x[3]))
plot(embed(washer3[1:N,], "Isomap", .mute = c("message", "output"), ndim =2, knn = 20)@data@data,
     xlab = "1st Dimension", ylab = "2nd Dimension",
     col = myColorRamp(c("red","purple","blue","green","yellow"), washer[,1])  )
```

### Rolled Washer

```{r}
N <- 2000;
par(mfrow = c(1,2))
scatterplot3d(washer.swiss[1:N,], color = myColorRamp(c("red","purple","blue","green","yellow"), washer[1:N,1] ),
               xlab = expression(x[1]), ylab = expression(x[2]), zlab = expression(x[3]))
plot(embed(washer.swiss[1:N,], "Isomap", .mute = c("message", "output"), ndim =2, knn = 20)@data@data,
     xlab = "1st Dimension", ylab = "2nd Dimension",
     col = myColorRamp(c("red","purple","blue","green","yellow"), washer[,1])  )
```

## Local Linear Embedding (LLE) {.tabset} 

Below, we show the manifold learning of the given manifolds using LLE with $K = 20$ nearest neighbors.

### Helix

```{r, message= FALSE}
par(mfrow = c(1,2) )
scatterplot3d(helix, color = myColorRamp(c("red","blue"), helix[,3]) )
plot(embed(helix, "LLE", .mute = c("message", "output"), ndim =1)@data@data, rep(0,dim(helix)[1]),
     xlab = "1st Dimension", ylab = "", yaxt = 'n',
     col = myColorRamp(c("red","blue"), helix[,3]) )
```

### Swiss Roll

```{r, message= FALSE}
par(mfrow = c(1,2))
scatterplot3d(Swiss, color = myColorRamp(c("red","purple","blue","green","yellow"), S ),
               xlab = expression(x[1]), ylab = expression(x[2]), zlab = expression(x[3]))
plot(embed(Swiss, "LLE", .mute = c("message", "output"), ndim =2, knn = 20)@data@data,
     xlab = "1st Dimension", ylab = "2nd Dimension",
     col = myColorRamp(c("red","purple","blue","green","yellow"), S)  )
```

### Folded Washer

```{r, message= FALSE}
N <- 2000;
par(mfrow = c(1,2))
scatterplot3d(washer3[1:N,], color = myColorRamp(c("red","purple","blue","green","yellow"), washer[1:N,1] ),
               xlab = expression(x[1]), ylab = expression(x[2]), zlab = expression(x[3]))
plot(embed(washer3[1:N,], "LLE", .mute = c("message", "output"), ndim =2, knn = 20)@data@data,
     xlab = "1st Dimension", ylab = "2nd Dimension",
     col = myColorRamp(c("red","purple","blue","green","yellow"), washer[,1])  )
```

### Rolled Washer

```{r, message= FALSE}
N <- 2000
par(mfrow = c(1,2))
scatterplot3d(washer.swiss[1:N,], color = myColorRamp(c("red","purple","blue","green","yellow"), washer[1:N,1] ),
               xlab = expression(x[1]), ylab = expression(x[2]), zlab = expression(x[3]))
plot(embed(washer.swiss[1:N,], "LLE", .mute = c("message", "output"), ndim =2, knn = 20)@data@data,
     xlab = "1st Dimension", ylab = "2nd Dimension",
     col = myColorRamp(c("red","purple","blue","green","yellow"), washer[,1])  )
```

