\documentclass[12 point]{article}
\usepackage[margin=1in]{geometry}
\usepackage{graphicx}
\usepackage{amsmath,amssymb}
\usepackage{amsthm}
\usepackage{array}
\usepackage{multirow}
\usepackage{tikz}
\usepackage{fancyhdr}
 \pagestyle{fancy}
\fancyhf{}
\rhead{Midterm}
\lhead{Statistics 185}
\lfoot{October 17, 2019}
\rfoot{Page \thepage}
\renewcommand{\headrulewidth}{1pt}
\renewcommand{\footrulewidth}{1pt}

\title{Statistics 185 - Midterm}
\date{October 17, 2019}
\begin{document}
\thispagestyle{empty}
\maketitle
\vspace{2cm}
\begin{center}Name(print):\line(1,0){300}\end{center}

\vspace{1.5cm}
\noindent This is a closed-book exam so do not refer to your notes or any
other books.  There are 6 problems.\\

\noindent Write your solutions as clearly as possible and make sure it's easy to find your answers (circle them if necessary), since you will not receive credit for work that I cannot understand or find. Good Luck! \\

\noindent If you find a question confusing, please ask me to clarify it. \\

\noindent If you need additional sheets, please ask and I will supply you with blank printer paper.

\vspace{1.25 cm}
\begin{center}
\renewcommand{\arraystretch}{1.5}
\begin{tabular}{| >{\centering\arraybackslash}m{2cm} | >{\centering\arraybackslash}m{2cm} | }
\hline
Question & Points\\
\hline
1 &  \\
\hline
2&  \\
\hline
3&  \\
\hline
4&   \\
\hline
5&   \\
\hline
6&   \\
\hline
\end{tabular} \\
\vspace{0.25cm}
\begin{tabular}{|>{\centering\arraybackslash}m{2cm} | >{\centering\arraybackslash}m{2cm} |}
\hline
Total & \\
\hline

\end{tabular}
\end{center}
\newpage
\begin{enumerate}
\item Suppose that $\vec{x}\in \mathbb{R}^p$ has mean $\vec{0}$ and covariance $I$.  Let $A \in \mathbb{R}^{q\times p}$ be a full rank matrix with $p > q$ and let  $\vec{y} = A\vec{x} + \vec{\epsilon}$ be a vector in $\mathbb{R}^q.$ where $\vec{\epsilon} \sim \mathcal{N}(0,\rho^2 I_q)$ is independent of $\vec{x}.$
\begin{itemize}
\item[a.] Find $\Sigma_y$ and $\Sigma_{xy}.$
    \vspace{8 cm}
\item[b.]  Find the canonical correlations between $\vec{x}$ and $\vec{y}.$ Express your answer in terms of $\rho^2$ and the singular values of $A$.
\end{itemize}
    
\newpage

\item Suppose $\vec{q}_1,\dots,\vec{q}_k$ are orthonormal vectors in $\mathbb{R}^d$ with $k < d.$  Suppose $$\vec{x} = a_1\vec{q}_1+\dots + a_k \vec{q}_k + \vec{\epsilon}$$
where $a_1,\dots,a_k$ are uncorrelated, mean zero random variables with variance $\lambda^2$ and $\vec{\epsilon}$ is independent of $a_1,\dots,a_k$ and follows a multivariate Normal distribution with mean $\vec{0}$ and covariance $\sigma^2 I$. Here, $I$ is the $d-$dimensional identity matrix.
\begin{itemize}
\item[a.] Find the mean of $\vec{x}$.
\vspace{3cm}
 \item[b.] Find the covariance matrix, $\Sigma$, of $\vec{x}$.
\vspace{5cm}
\item[c.] Give the eigenvalues and respective eigenvectors of $\Sigma.$  (Hint: you may assume there are orthonormal vectors $\vec{q}_{k+1},\dots,\vec{q}_d$ which are a basis for the orthogonal complement of the span of $\vec{q}_1,\dots,\vec{q}_d.$)
\end{itemize}
\newpage
\item Let $\vec{x}_1,\dots,\vec{x}_N$ be vectors in $\mathbb{R}^d$. Assume a PCA of these data has loadings $\vec{w}_1,\dots,\vec{w}_d$ with associated variances $\lambda_1 \ge \dots \ge \lambda_d \ge 0.$ Let $U$ be a $d\times d$ orthonormal matrix and set $\vec{y}_i = U\vec{x}_i.$ Find expressions for the principal components loadings and variance of $\vec{y}_1,\dots,\vec{y}_N$ in terms of $U$, $\vec{w}_1,\dots,\vec{w}_d$ and $\lambda_1,\dots,\lambda_d.$


\newpage
\item Let $$ \| X - WH\|_F = \sqrt{\sum_{i,j}[X_{ij} - (WH)_{ij}]^2}$$ where $X \in \mathbb{R}^{N\times p}$, $W\in \mathbb{R}^{N\times k}$, and $H\in \mathbb{R}^{k\times p}$.
\begin{itemize}
\item[a.] Find the partial derivative of $\frac{1}{2}\| X - WH\|_F^2$ with respect to $W_{ij}$.
\vspace{9 cm}
\item[b.] Find the partial derivative of $\frac{1}{2}\| X - WH\|_F^2$ with respect to $H_{jl}$.
\end{itemize}

\newpage
\item Suppose $\vec{x} \in \mathbb{R}^p$ and $\vec{y} \in \mathbb{R}^q$ are random vectors with covariance matrices $\Sigma_x$ and $\Sigma_y$ respectively.  Suppose $\vec{x}$ and $\vec{y}$ have cross-covariance matrix $\Sigma_{xy}$ with rank $k \le \min \{p,q\}$ and canonical correlation variable $\eta_1,\dots,\eta_k$ and $\xi_1,\dots,\xi_k$ with canonical correlations $\sigma_1\ge \dots \ge \sigma_k>0$.  
	\begin{itemize}
 	 \item[a.]  Find the covariance matrices of the random vectors $\vec{\eta} = 					(\eta_1,\dots,\eta_k)^T$ and $\vec{\xi} = (\xi_1,\dots,\xi_k)^T.$
    		\vspace{8 cm}
   	 \item[b.]  Find the cross-covariance of $\vec{\eta}$ and $\vec{\xi}.$
    \end{itemize}


\newpage
\item Let $\vec{x}_1,\dots,\vec{x}_N$ be random vectors in $\mathbb{R}^p$ with sample covariance $\hat{\Sigma}_X$ and $\vec{y}_1,\dots,\vec{y}_N$ be random vectors in $\mathbb{R}^q$ with covariance matrix $\hat{\Sigma}_Y$.  Assume that $\hat{\Sigma}_X$ and $\hat{\Sigma}_Y$ are both full rank, i.e. the principal component variances of both $\vec{x}_1,\dots,\vec{x}_n$ and the principal component variances of $\vec{y}_1,\dots,\vec{y}_N$ are all positive.  

Let $\vec{z}_i ^T = (\vec{x}_i^T,\vec{y}_i^T)$ for $i=1,\dots,N$ be random vectors in $\mathbb{R}^{p+q}$ with sample covariance $\hat{\Sigma}_Z.$  Now assume $\hat{\Sigma}_Z$ is not full rank, i.e. at least one of the principal component variances of $\vec{z}_1,\dots,\vec{z}_N$ is zero. Is the largest canonical correlation between $\{\vec{x}_i\}_{i=1}^N$ and $\{\vec{y}_i\}_{i=1}^N$ equal to 1?  Explain your answer.  

Hint: If $A\in\mathbb{R}^{p \times p}$, $B\in\mathbb{R}^{p\times q}$, $C\in \mathbb{R}^{q\times p}$, and $D\in\mathbb{R}^{q\times q}$ are matrices and $A$ is invertible then
$$det\bigg(\begin{bmatrix} A&B\\C& D\end{bmatrix} \bigg) = det(A) det(D - CA^{-1}B).$$  You do not need this hint. It just offers one route to a solution.

\end{enumerate}


\end{document}