--- 
title: 'Homework #1'
author: 
output:
  html_document:
    df_print: paged
---

# {.tabset}
This assignment is a review of important topics from linear algebra.  Additionally, it provides an introduction to R, RStudio, and RMarkdown for new users.  Be sure to uncomment (remove the `#` from the R code) before running the commands.


## Linear Algebra Review 

This assignment should cover a high level review of undergraduate linear algebra.  Topics such as spans, subspaces and dimension, basis, rank, eigenvectors/values.  


## R Coding 

For many undergraduates R is unfamiliar.  Sample problems below are intended to refresh or introduce R which is used extensively throughout the course.

1) Consider the matrix 

a. Write R code to add $A$ as a variable in R. 
\[
A = \left[  
\begin{matrix}
1 & 0 & 1 & 0 \\
0 & 1 & 0 & 0 \\
1 & 0 & 1 & 0\\
0 & 0 & 0 & 1
\end{matrix}
\right]
\]
    
```{r, collapse=TRUE}
# A <- 
```

b. Find the eigenvalues and corresponding eigenvectors of $A$. Write these to a variable called *A.eigen*. You may find the command *eigen()* helpful. 

```{r, collapse=TRUE}
# A.eigen <- 
```
    
c. Find matrices $P$ and $D$ such that $A=PDP^{-1}$ and write these to variables in R.

```{r, collapse=TRUE}
# P <- 
# D <-
```
    
    
d. Let's calculate $PDP^{-1}$ in R to verify your answer to part (c) is correct.

```{r, collapse=TRUE}
# A - P %*% D %*% solve(P)
```
    
2. Consider the `mtcars` dataset which is available from the `datasets` library. We can load the `datasets` library and view the `mtcars` dataset using the chunk of code: 

```{r, collapse = TRUE}
library(datasets) # load the datasets 
mtcars # view the mtcars dataset
```
We can also view a summary of the *mtcars* dataset using the command  *summary*:
```{r}
summary(mtcars) # view a summary of the mtcars dataset
```
Write commands to complete the following tasks.

a. List the names of all cars with 4 gears.
```{r}
# your code here
```
    
b. List the names of all cars with at least 100 horsepower.
```{r}
# your code here
```
    
c. Find the highest mpg.
```{r}
# your code here
```
    
d. Find the name of the car with the highest mpg.
```{r}
# your code here
```
  
e. Create a scatterplot of the *mpg* (dependent variable) vs. *wt* (independent variable).
```{r}
# your code here
```

